sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-01-28 doc/paper.Rmd
Copy edit end of paper
Edit/ move figs/tables around

<!-- TODO: rotate and fill page: -->

@@

<!-- Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models. -->
We can get a summary of the fit with the `print()` method or extract parameter estimates as a data frame with the `tidy()` method.
<!-- (omitted for space). -->

```{r,eval=FALSE, echo=TRUE}
print(fit)
```

@@

We could extend this spatial model to be a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured in time.
Here we will turn off the spatial field and structure the spatiotemporal random fields as AR(1).

@@

(ref:timing-fig) Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
Lines represent means and ribbons ranges across 20 random iterations.
Note the log x and y axes.
sdmTMB used the default settings and `normalize = TRUE`, which calculates the random effect normalizing constant in R.
VAST should be similar to sdmTMB and so is not shown.
inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
mgcv used `bam()`, `method = "fREML"`, and discretized covariates [@miller2019].
<!-- inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library. -->
<!-- All platforms were restricted to one core and could be faster with parallel computation. -->
<!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->

@@

We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \@ref(fig:timing), Appendix 6).
In this test, for mesh resolutions under 500 nodes, sdmTMB was faster than inlabru or mgcv.
inlabru was the least affected by mesh resolution and beyond 500 nodes was faster than sdmTMB with default sdmTMB settings.
mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown).
<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
Our test was restricted to one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speed increase for mgcv.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that VAST does not) -->

@@

sdmTMB would not be possible without the TMB [@kristensen2016] and INLA [@rue2009; @lindgren2011; @lindgren2015] R packages.
sdmTMB is heavily inspired by and in some places code has been adapted from both the VAST [@thorson2019] and glmmTMB [@brooks2017] R packages.
Penalized spline support was possible thanks to mgcv [@wood2017a].
We thank the authors of all these packages.
We thank M. Lindmark, P.M. Regular, J.T. Thorson, J. Watson, ..., TODO.

@@

SA led development of the software and EW contributed key software code. All
authors contributed software code, guided software development, and
participated in testing. PE led the Snowy Owl case study. All authors
contributed to appendix code. SA led the writing of the manuscript. All authors
contributed critically to the drafts and gave final approval for publication.

>>>>

@@

<!-- Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models. -->
We can get a summary of the fit with the `print()` or `summary()` methods or extract parameter estimates as a data frame with the `tidy()` method.
<!-- (omitted for space). -->

```{r,eval=FALSE, echo=TRUE}
summary(fit)
```

@@

We could extend this spatial model to be a spatiotemporal one simply by supplying the year column name to the `time` argument and specifying how we want the spatiotemporal random fields to be structured.
Here we will keep the spatial field and structure the spatiotemporal random fields as AR(1).

@@

We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \@ref(fig:timing), Appendix 6).
In this test, for mesh resolutions under 500 nodes, sdmTMB was faster than inlabru or mgcv.
inlabru was the least affected by mesh resolution and beyond 500 nodes was faster than sdmTMB with default sdmTMB settings.
mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
Our test was restricted to one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speed increase for mgcv.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that VAST does not) -->

(ref:timing-fig) Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
Lines represent means and ribbons ranges across 20 random iterations.
Note the log x and y axes.
sdmTMB used the default settings and `normalize = TRUE`, which calculates the random effect normalizing constant in R.
VAST is similar to sdmTMB and so is not shown.
inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
mgcv used `bam()`, `method = "fREML"`, and discretized covariates [@miller2019].
<!-- inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library. -->
<!-- All platforms were restricted to one core and could be faster with parallel computation. -->
<!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->

@@

@@

sdmTMB would not be possible without the TMB [@kristensen2016] and INLA [@rue2009; @lindgren2011; @lindgren2015] R packages.
sdmTMB is heavily inspired by and in some places code has been adapted from both the VAST [@thorson2019] and glmmTMB [@brooks2017] R packages.
Penalized spline support was possible thanks to mgcv [@wood2017a].
We thank the authors of all these packages.
We thank M. Lindmark, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.

@@

SA led development of the software and EW contributed key software code.
All authors contributed software code, guided software development, and
participated in testing.
PE led the Snowy Owl case study.
All authors contributed to appendix code.
SA led the writing of the manuscript; all authors contributed critically to the drafts and gave final approval for publication.
